The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-09-15 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Tada
Stats... etc

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[font={it},labelfont={bf}]{caption}
%\usepackage{units}
\usepackage{times}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{xspace}
%\usepackage{subcaption}

@@

\section{Method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern -- in particular for a many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). LOFAR's \textsc{aoflagger} has shown to have a good accuracy, is fast and has a library interface, which allows it to be integrated in a pipeline. It is therefore also used for MWA's RFI detection.

@@

\begin{table*}
\caption{List of observations used in the analyses. The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives). The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag. Because the DVB frequencies are only observed in GLEAM observations, this fraction is only given for those. }\label{tbl:obs-list}
\begin{tabular}{|l|l|r|r|r|c|}
\hline
\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
\hline
EoR   & 2013-08-23 & 138.8--197.7              & 3h & 0.53\% & ---\\
GLEAM & 2013-08-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.94\% & 0\% \\
EoR   & 2013-08-26 & 138.8--197.7              & 6h & 0.81\% & ---\\
GLEAM & 2013-11-05 & 72.3--133.8, 138.8--230.8 & 7h & 1.30\% & 0\% \\
GLEAM & 2013-11-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.69\% & 0\% \\
EoR   & 2014-02-05 & 138.8--197.7              & 6h & 0.54\% & -- \\
GLEAM & 2014-03-16 & 72.3--133.8, 138.8--230.8 & 7h & 0.79\% & 0\% \\
GLEAM & 2014-03-17 & 72.3--133.8, 138.8--230.8 & 7h & 1.64\% & 3\% \\
EoR   & 2014-04-10 & 138.8--197.7              & 6h & 0.68\% & ---\\
GLEAM & 2014-06-18 & 72.3--133.8, 138.8--230.8 & 7h & 0.98\% & 0\% \\
RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
\hline
\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.97\%} & \textbf{0.5\%}\\
\hline
\end{tabular}
\end{table*}

@@

The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2. Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz. In these observations, the selected EoR field is tracked byt the primary beam, such that the sensitivity towards the field is maximal.

@@

Both projects avoid the subbands that cover the frequency range 133.8--138.8~MHz, because the ORBCOMM low Earth orbitting (LEO) satellites transmit in these subbands. When these frequencies are observed, the subbands are often so strongly contaminated that it affects imaging sensitivity. For completeness, we include an 16-min observation that covers this frequency range.

@@

\noindent\begin{figure*}
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
\caption{Strong RFI contamination found in the 2 m amateur band (146 MHz), which falls in the lower band of the MWA EoR observations. These panels show the Stokes I amplitudes. In the right panel, the result of RFI detection is shown with purple. The thin horizontal lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively. Two more transmitters are detected in these data: at 150.17 and 156.7 Mhz, which are both one 40 kHz channel wide. These transmitters are not distinguishable in this plot, but are shown in the next figures.}
\label{fig:amateur-2m}
\end{center}
\end{figure*}

\noindent\begin{figure*}
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/150_2_mhz_example}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged}
\caption{Short RFI burst of 2 s centred on 150.17 MHz. This shows that flagging at high time and frequency resolution is necessary to accurately detect RFI.}
\label{fig:150_2}
\end{center}
\end{figure*}

\noindent\begin{figure*}
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
\caption{A weakly-observed transmitter centred on 156.66 MHz. Although it is hard to identify in such a time-frequency plot, the same frequency is flagged in several correlations, and it can therefore be assumed this is real RFI. This is an example in which the AOFlagger detect an event that would not be detected by regular thresholding. }
\label{fig:156_7}
\end{center}
\end{figure*}

@@

>>>>

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[font={it},labelfont={bf}]{caption}
%\usepackage{units}
\usepackage{times}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{xspace}
\usepackage{subcaption}

@@

\section{Method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern -- in particular for a many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). LOFAR's \textsc{aoflagger} has shown to have a good accuracy, is fast and has a library interface, which allows it to be integrated in a pipeline. It is therefore used for MWA's RFI detection.

@@

\begin{table*}
\caption{List of observations used in the analyses. The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives). The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag. Because the DVB frequencies are only observed in GLEAM observations, this fraction is only given for those. }\label{tbl:obs-list}
\begin{tabular}{|l|l|r|r|r|c|}
\hline
\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
\hline
EoR   & 2013-08-23 & 138.8--197.7              & 3h & 0.53\% & ---\\
GLEAM & 2013-08-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.94\% & 0\% \\
EoR   & 2013-08-26 & 138.8--197.7              & 6h & 0.81\% & ---\\
GLEAM & 2013-11-05 & 72.3--133.8, 138.8--230.8 & 7h & 1.27\% & 0\% \\
GLEAM & 2013-11-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.69\% & 0\% \\
EoR   & 2014-02-05 & 138.8--197.7              & 6h & 0.54\% & -- \\
GLEAM & 2014-03-16 & 72.3--133.8, 138.8--230.8 & 7h & 0.79\% & 0\% \\
GLEAM & 2014-03-17 & 72.3--133.8, 138.8--230.8 & 7h & 1.64\% & 3\% \\
EoR   & 2014-04-10 & 138.8--197.7              & 6h & 0.68\% & ---\\
GLEAM & 2014-06-18 & 72.3--133.8, 138.8--230.8 & 7h & 0.98\% & 0\% \\
RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
\hline
\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{...\%} & \textbf{0.5\%}\\
\hline
\end{tabular}
\end{table*}

@@

The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2. Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz. In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal.

@@

Both projects avoid the subbands that cover the frequency range 133.8--138.9~MHz, because the ORBCOMM low Earth orbitting (LEO) satellites transmit in these subbands. When these frequencies are observed, the subbands are often so strongly contaminated that it affects imaging sensitivity. For completeness, we include an 16-min observation that covers this frequency range.

@@

\noindent\begin{figure*}
\subcaptionbox{RFI contamination found in the 2 m amateur band (146 MHz).}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
\label{fig:amateur-2m}
}\\\vspace{3mm}%
\subcaptionbox{Short RFI burst of 2 s centred on 150.17 MHz. This shows that flagging at high time and frequency resolution is necessary to accurately detect RFI.}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/150_2_mhz_example}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged}
\label{fig:150_2}
}\\\vspace{3mm}%
\subcaptionbox{A weakly-observed transmitter centred on 156.66 MHz. This same frequency is flagged in several correlations, and it can therefore be assumed this is real RFI. This is an example in which the AOFlagger detects an event that would not be detected by regular thresholding. }[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
\label{fig:156_7}
}%
\caption{Examples of RFI events found in the EoR low band (138.9--197.7 MHz) in a 2-min snapshot with relatively high RFI contamination. These panels show the Stokes I amplitudes. In the right figures, the result of RFI detection is shown with purple. The horizontally flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
\end{figure*}

@@

\noindent\begin{figure}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/rfi-per-band}
\caption{RFI per subband}
\label{fig:rfi-per-band}
\end{center}
\end{figure}
